Embeddings and Attention in Predictive Modeling

### 2022-03-31 manuscript/manuscript.Rmd
Updates for review

---
title: Embeddings and Attention in Predictive Modeling 
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevinykuo@gmail.com 
  - name: Ronald Richman
    affiliation: QED Actuaries and Consultants and University of the Witwatersrand
    email: ron@ronaldrichman.co.za
abstract: |
  We explore in depth how categorical data can be
  processed with embeddings in the context of claim severity modeling. We develop several models that range
  in complexity from simple neural networks to state-of-the-art attention based architectures that utilize embeddings. 
  We illustrate the utility of learned embeddings from neural networks as pretrained features in
  generalized linear models, and discuss methods for visualizing and interpreting embeddings. Finally, we explore
  how attention based models can contextually augment embeddings, leading to enhanced predictive performance.
keywords:
  - categorical data
  - neural networks
  - embeddings
  - attention
  - severity modeling
  - National Flood Insurance Program 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes: |
    \usepackage{amsmath}
    \usepackage{mathtools}
    \usepackage{bm}
    \usepackage{bbm}
    \usepackage{float}
    \DeclarePairedDelimiter\ceil{\lceil}{\rceil}
    \newcommand{\norm}[1]{\| #1 \|}
---

@@

We note that the focus of our contribution is on applications of categorical embeddings in predictive modeling, rather than on fine-tuning models in order to beat or establish a benchmark. To this end, we do not perform systematic hyperparameter tuning in our experiments, and the performance metrics are provided to show that reasonable models utilizing categorical embeddings can be developed. However, as we describe the modeling procedure and results, we provide commentary on recommended practices in practical applications, especially with respect to the neural network models.

@@

\begin{equation}
1 + |J^{\text{num}}| + \left(\sum_{j \in J^{\text{cat}}} n_{\mathcal{P}}^j\right) - |J^{\text{cat}}|:
(\#eq:glm1)
\end{equation}

one for the intercept, one for each numeric variable, and the numbers of levels minus one (due to dummy coding) for each categorical variable.

@@

For Model 2, we build a feedforward neural network, whose architecture is shown in Figure \@ref(fig:nn-1). The categorical inputs go through one-dimensional embedding layers where they are each mapped to a scalar. The embeddings are then concatenated with the numeric predictors, which have been normalized in data pre-processing, before being passed through a feedforward layer (with 8 hidden units and ReLU activation) to obtain a scalar value between 0 and 1, as constrained by a sigmoid output activation. This scalar output represents the proportion of coverage amount that is paid out, which we then multiply by the dollar amount of the coverage amount to obtain our prediction of the claims paid. For optimization, we use the mean squared error (MSE) loss function.

@@

For Model 4, the neural network with multi-dimensional embeddings, we would need to treat the learned values before we can plot them. Specifically, we need to reduce the dimension of the data to 2D for embeddings with more than two dimensions. The standard approaches for this are principal component analysis (PCA; @shlens2014tutorial) and t-distributed stochastic neighbor embedding (t-SNE; @Maaten2008a), both of which we illustrate. For a technical review of these methods, refer to Appendix \ref{dim-red].

@@

```{r numdisc, echo = FALSE, fig.cap = "Distribution of the Comunity Rating System Discount covariate", out.width = "60%", fig.align = "center"}
knitr::include_graphics("images/community_rating_system_discount.pdf")
```

@@

\begin{equation}
p_{ij} = \frac{p_{j|i} + p_{i|j}}{2n}.
\end{equation}

On the other hand, in the target space, t-SNE assumes the neighbor selection is driven by a Student t-distribution with one degree of freedom, i.e., a Cauchy distribution. The similarity metric for a pair of points in $\mathcal{Y}$ is then defined as the joint probability

\begin{equation}
q_{ij} = \frac{(1 + \norm{y_i - y_j}^2)^{-1}}{\sum_{k \neq l} (1 + \norm{y_k - y_l}^2)^{-1}}.
\end{equation}

With the similarity metrics defined for each space, we can then state the objective for "closeness" for the sets of points, which is the Kullback--Leibler divergence from Q to P:

\begin{equation}
D_{KL}(P || Q) = \sum_{i \neq j} p_{ij} \log \frac{p_{ij}}{q_{ij}}. (\#eq:tsne-obj)
\end{equation}

We now circle back to Equation \@ref(eq:pji) and discuss $\sigma_i$. The variance of the Gaussian is allowed to vary depending on where the point is in the space; specifically, it should reflect the density or sparsness of surrounding points. To train this parameter, t-SNE introduces a hyperparameter, *perplexity*, which is a scalar value set by the user, and is defined as

@@

\begin{equation}
H(P_i) = -\sum_j p_{j|i}\log_2 p_{j|i},
\end{equation}

and $P_i$ is the Gaussian parameterized by $x_i$ and $\sigma_i$. During training, t-SNE looks for $P_i$ that have perpelxity corresponding to the user input. Finally, the objective as stated in Equation \@ref(eq:tsne-obj) is differentiable, and is trained via gradient descent.

>>>>

---
title: Embeddings and Attention in Predictive Modeling 
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevinykuo@gmail.com 
  - name: Ronald Richman
    affiliation: Old Mutual Insure and University of the Witwatersrand
    email: ron@ronaldrichman.co.za
abstract: |
  We explore in depth how categorical data can be
  processed with embeddings in the context of claim severity modeling. We develop several models that range
  in complexity from simple neural networks to state-of-the-art attention based architectures that utilize embeddings. 
  We illustrate the utility of learned embeddings from neural networks as pretrained features in
  generalized linear models, and discuss methods for visualizing and interpreting embeddings. Finally, we explore
  how attention based models can contextually augment embeddings, leading to enhanced predictive performance.
keywords:
  - categorical data
  - neural networks
  - embeddings
  - attention
  - severity modeling
  - National Flood Insurance Program 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes: |
    \usepackage{amsmath}
    \usepackage{mathtools}
    \usepackage{bm}
    \usepackage{bbm}
    \usepackage{float}
    \DeclarePairedDelimiter\ceil{\lceil}{\rceil}
    \newcommand{\norm}[1]{\| #1 \|}
---

@@

\textcolor{blue}{Whereas the main focus of our contribution} is on applications of categorical embeddings in predictive modeling, \textcolor{blue}{we have fine-tuned the neural network models in this (and the next) sections using the same 5-fold cross validation scheme described above, but applied to a separate 100k data point subsample of the FEMA data. The results of this fine-tuning are reported in detail in the appendix. The performance metrics in this, and the next, section are based on the optimal models found through this process. Our aim is thos show that reasonable models utilizing categorical embeddings can be developed quite easily. As we describe the modeling procedure and results, we provide further commentary on recommended practices in practical applications, especially with respect to the neural network models.}

@@

```{=tex}
\begin{equation}
1 + |J^{\text{num}}| + \left(\sum_{j \in J^{\text{cat}}} n_{\mathcal{P}}^j\right) - |J^{\text{cat}}|:
(\#eq:glm1)
\end{equation}
```
one for the intercept, one for each numeric variable, and the numbers of levels minus one (due to dummy coding) for each categorical variable.

@@

For Model 2, we build a feedforward neural network, whose architecture is shown in Figure \@ref(fig:nn-1). The categorical inputs go through one-dimensional embedding layers where they are each mapped to a scalar. The embeddings are then concatenated with the numeric predictors, which have been normalized in data pre-processing, before being passed through a feedforward layer (with 8 hidden units and ReLU activation) to \textcolor{blue}{finally} obtain a scalar value between 0 and 1, as constrained by a sigmoid output activation. \textcolor{blue}{Some regularization is applied within this network using the dropout technique of} @srivastava2014dropout \textcolor{blue}{and by early-stopping the fitting of the network by selecting the model weights that relate to the lowest estimated MSE score on a 5\% validation sample.} This scalar output represents the proportion of coverage amount that is paid out, which we then multiply by the dollar amount of the coverage amount to obtain our prediction of the claims paid. For optimization, we use the mean squared error (MSE) loss function.

@@

For Model 4, the neural network with multi-dimensional embeddings, we would need to treat the learned values before we can plot them. Specifically, we need to reduce the dimension of the data to 2D for embeddings with more than two dimensions. The standard approaches for this are principal component analysis (PCA; @shlens2014tutorial) and t-distributed stochastic neighbor embedding (t-SNE; @Maaten2008a), both of which we illustrate. For a technical review of these methods, refer to Appendix \\ref{dim-red].

@@

```{r numdisc, echo = FALSE, fig.cap = "Distribution of the Comunity Rating System Discount covariate", out.width = "60%", fig.align = "center"}
knitr::include_graphics("images/community_rating_system_discount.pdf")
```
# Finetuning the Neural Network Models
Here we describe how the neural network models presented in Sections 4 and 5 above were fine-tuned.

## Models in Section 4
\textcolor{blue}{
As noted before, a different 100k subsample of the NFIP dataset was produced for this purpose, and a 5-fold cross-validation scheme was implemented using this data. Two main hyperparameters were tuned for each of Model 2 and Model 4. Two different activation functions (the ReLu and hyperbolic tangent activation functions) and 5 different widths of the hidden layers (8, 16, 32, 64, 128) were tested. The following table shows these results for each of Model 2 and Model 4.}

|  | ReLU/8 | ReLu/16 | ReLu/32 | ReLu/64 | ReLu/128 | tanh/8 | tanh/16 | tanh/32 | tanh/64 | tanh/128 |
|---|---|---|---|---|---|---|---|---|---|---|
| Model 2 - RMSE |          60 988  |          60 561  |          60 125  |          60 147  |          59 918  |          61 728  |          60 506  |          60 130  |          59 893  |          60 065  |
| Model 4 - RMSE |          60 535  |          60 724  |          60 311  |          60 191  |          60 114  |          60 521  |          60 658  |          60 279  |          59 892  |          59 987  |
| Model 2 - MAE |          36 604  |          36 791  |          36 936  |          37 140  |          36 816  |          37 206  |          36 759  |          36 921  |          37 103  |          37 177  |
| Model 4 - MAE |          36 828  |          36 824  |          36 780  |          37 054  |          36 672  |          36 528  |          36 725  |          36 483  |          36 766  |          36 394  |
: Results of fine-tuning models in Section 4.

\textcolor{blue}{It can be observed that the lowest RMSE values are produced using the tanh activation function with a hidden layer width 64 neurons.}

## Models in Section 5

\textcolor{blue}{Models 5 and 6 add an attention-based mechanism to derive contextual embeddings to the models in Section 4. To find optimal hyperparameters, the best performing variation of Model 4 (tanh/64) was taken as a baseline, and the attention mechanisms described for Models 5 and 6 were added to this baseline. Since the performance of the models would ideally improve as a result of adding the attention mechanisms, and, indeed, we see this is the case in the following table, we do not attempt to fine-tune Models 5 and 6 any further.}

@@

```{=tex}
\begin{equation}
p_{ij} = \frac{p_{j|i} + p_{i|j}}{2n}.
\end{equation}
```
On the other hand, in the target space, t-SNE assumes the neighbor selection is driven by a Student t-distribution with one degree of freedom, i.e., a Cauchy distribution. The similarity metric for a pair of points in $\mathcal{Y}$ is then defined as the joint probability

```{=tex}
\begin{equation}
q_{ij} = \frac{(1 + \norm{y_i - y_j}^2)^{-1}}{\sum_{k \neq l} (1 + \norm{y_k - y_l}^2)^{-1}}.
\end{equation}
```
With the similarity metrics defined for each space, we can then state the objective for "closeness" for the sets of points, which is the Kullback--Leibler divergence from Q to P:

```{=tex}
\begin{equation}
D_{KL}(P || Q) = \sum_{i \neq j} p_{ij} \log \frac{p_{ij}}{q_{ij}}. (\#eq:tsne-obj)
\end{equation}
```
We now circle back to Equation \@ref(eq:pji) and discuss $\sigma_i$. The variance of the Gaussian is allowed to vary depending on where the point is in the space; specifically, it should reflect the density or sparsness of surrounding points. To train this parameter, t-SNE introduces a hyperparameter, *perplexity*, which is a scalar value set by the user, and is defined as

@@

```{=tex}
\begin{equation}
H(P_i) = -\sum_j p_{j|i}\log_2 p_{j|i},
\end{equation}
```
and $P_i$ is the Gaussian parameterized by $x_i$ and $\sigma_i$. During training, t-SNE looks for $P_i$ that have perpelxity corresponding to the user input. Finally, the objective as stated in Equation \@ref(eq:tsne-obj) is differentiable, and is trained via gradient descent.
